---
title: sdmTMB setup on DFO computers
---

1. Install the following programs (on DFO Windows laptops, using the
   'Software Centre'):

- R 4.5.1 (or 4.4 or higher)
- RStudio latest version (for DFO computers you need to search
  "R Studio" with a space)

2. In R, install sdmTMB from CRAN

```{r, eval=FALSE}
install.packages("sdmTMB", dependencies = TRUE)
```

- If asked if you'd like to update any packages, say yes.
- If CRAN asks if you'd like to build a newer package from source,
  feel free to say no.
- Sometimes some packages fail to install on DFO laptops. If any
  packages fail (e.g. 'non-zero exist status), try installing them
  separately and then trying to install sdmTMB again.

If you get a warning like this:

```
TMB was built with Matrix version 1.5.3
Current Matrix version is 1.5.1
Please re-install 'TMB' from source using
install.packages('TMB', type = 'source') or ask
CRAN for a binary version of 'TMB' matching CRAN's
'Matrix' package
```

You can safely ignore it (if you want to squash the warning, you can
install TMB from source as described).

3. Test to make sure everything is working:

```{r, eval=FALSE}
library(sdmTMB)
fit <- sdmTMB(
  density ~ s(depth),
  data = pcod_2011, 
  mesh = pcod_mesh_2011,
  family = tweedie(link = "log")
)
fit
```

If successful, you should see this:

```
Spatial model fit by ML ['sdmTMB']
Formula: density ~ s(depth)
Mesh: pcod_mesh_2011 (isotropic covariance)
Data: pcod_2011
Family: tweedie(link = 'log')
 
Conditional model:
            coef.est coef.se
(Intercept)     2.16    0.34
sdepth          1.94    3.13

Smooth terms:
             Std. Dev.
sd__s(depth)     13.07

Dispersion parameter: 13.68
Tweedie p: 1.58
Matérn range: 16.84
Spatial SD: 2.20
ML criterion at convergence: 2937.789

See ?tidy.sdmTMB to extract these values as a data frame.
```

Again, you can ignore any warnings about Matrix/TMB versions if you see those.

## Other packages

```{r}
pkgs <- c("dplyr", "ggplot2", "sf", "inlabru", "purrr",
  "future", "here", "knitr", "rnaturalearth", "fishMod")
install.packages(pkgs)
```

## Optionally installing the development version of sdmTMB

This is completely optional. If this is confusing or doesn't work,
don't worry---the CRAN version is just fine!

This requires a C++ compiler or the Rtools application on Windows.
Install Rtools 4.5 from the Software Center.

```{r, eval=FALSE}
pak::pak("pbs-assess/sdmTMB", dependencies = TRUE)

# if that fails, you can also try:
# remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
```
